Numerical trajectory patterns of floating macroalgae

### 2020-10-29 thesis_chapter_3.Rmd
updated rmd and outputs. table added

---
title: 'Numerical trajectory patterns of floating macroalgae'
author: "Ross Coppin"
output:
  html_document:
  pdf_document:
    fig_caption: yes
  word_document:
    reference_docx: markdown/thesis_template.docx
bibliography: references.bib
csl: markdown/frontiers.csl.txt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

@@

To accurately determine the trajectory of marine macroalgae these aspects need to be taken into account. Past research by @allen1999 and @breivik2011 have provided estimates of drag for various objects based on experimental work, which consisted of both direct and indirect methods of the Leeway model. However, this work does not consider biological material such as macroalgae. Although drag estimates do exist, these have been calculated for macroalgae not detached from the substratum and are regarded as fixed-point estimates. Methods used in the field of chemical engineering provide an alternative approach to estimating drag of macroalgae by means of formulas to calculate drag based on shape. Chemical engineers in the field of fluid dynamics which come across similar problems when including drag use an approach known as a 'spherecity factor'. A spherecity factor is the ratio of the surface area of a sphere having the volume as the irregular shaped particle (s) to the actual surface area of the particle (s_d) such that $$\psi=s/s_d$$. The values are dependent on the shape of the object and a range of values have been 
determined based on numerous experimental work. A table of various shapes and the associated spherecity factor is provided in the appendix.

There are considerable gaps in the literature with regards to macroalgae trajectory such as buoyancy, air form drag, water form drag, epibiont load, growth, fragmentation, size, shape and orientation which have not been investigated numerically in the past. In this study, a hypothetical approach was used to shed light on the role of drag on macroalgal drift trajectory by means of Langrangian based trajectory simulations. These simulations aimed to determine the role that hydrodynamic and wind drag might play in drift trajectory of macroalgae. In terms of wind drag, the direct effect of wind on the kelp was estimated rather than its affect on surface-currents.

@@

## Load software:
library(ggplot2)
library(rgdal)
library(tidyverse)
library(viridis)
library(gridExtra)
source("map/scaleBarFunc.R")

@@

library(ggpubr)

@@

This study simulated two scenarios; one where the plant is fully submerged and only exposed to seawater drag; and the other where parts of the plants are partially exposed resulting in a combination of seawater and air form drag. Drag based on shape based coefficients were used in the calculations for determining the overall kelp velocity vectors. A momentum energy equation was used to calculate the drag forces for the relevant simulation. This approach has been employed in modeling iceberg drift trajectory [@andersson2017; @lichey2001; @eik2009] and was adapted to suit this particular study. The energy momentum equation used to calculate the drag force exerted on the virtual kelp particle was, $$m({d_u/d_t})=F_d + F_a + F_f$$
where $m$ is kelp mass, $d_u$ the virtual kelp particle velocity, $F_d$ the hydrodynamic drag, $F_a$ is the wind drag, and $F_f$ the surface-current flow. The assumption made was that the mass of the kelp did not change for each virtual kelp particle over the course of the simulations. Hydrodynamic drag, wind drag and surface current-flow are two-dimensional vector quantities and were calculatedfor each time step of each simulation.

@@

In order to incorporate hydrodynamic and wind drag, the cross-sectional area of the kelp was calculated first. Known geometric shapes reflecting the relevant plant sections were used to estimate the surface area for various parts of the plant, for details please see image/table ???. The dimensional data needed was estimated in cases were data was not available for that particular morphological characteristic. The bulb/pneumatocyst is a highly variable morphological characteristic and in some cases can appear absent, the same is true for the holdfast. Therefore, the cross-sectional area of the bulb/pneumatocyst was not considered in the calculation of overall cross-sectional area. In terms of the holdfast, a standard cross-sectional area was used for simulations where the inclusion of the holdfast was considered.

The cross-sectional area calculated was site specific and the dimensions needed were garnered from morphology data from a previous study [@coppin2020]. The minimum, maximum and mean were calculated for each morphological characteristic (see appendix), which were used for calculating the overall minimum, maximum and mean cross-sectional areas needed to run the various simulations. The trajectory of these "types" of cross-sectional areas (minimum, maximum and mean overall cross-sectional area) were used to determine the influence of drag, both hydrodynamic and wind, on overall trajectory.

@@

```{r morphology_wrangling, message=FALSE, warning=FALSE, include=FALSE}
# Investigate morphology data to assist in estimating surface area

# Load software
library(tidyverse)
library(ggpubr)
library(gridExtra)
library(ggplot2)
library(ggord)
library(vegan)
library(dplyr)

# Read in morphology data
eck_df <- read.csv("data/morph_data/eck_morph_data.csv", sep = ";") %>%
  filter(site!= "Rocky Bank") %>%   # Not using Rocky Bank in this study
  filter(site!= "De Hoop")

eck_df$site <- as.character(eck_df$site)
eck_df <- eck_df %>% 
  filter(population == "Deep")

eck_long_df <- gather(eck_df, Morphology, measurement, frond_mass:total_length, factor_key=TRUE)
# Summary stats to make it easier to refer to later, these will not be used any analysis.

site_list <- c("Kommetjie")

eck_stats <- eck_long_df %>% 
  group_by(site, Morphology) %>% 
  summarise(mean = mean(measurement, na.rm = T),
            minimum = min(measurement, na.rm = T),
            maximum = max(measurement, na.rm = T)) %>% 
  ungroup() %>% 
  filter(site %in% site_list) %>% 
  rename(Site = site) %>% 
  rename(Mean = mean) %>% 
  rename(Minimum = minimum) %>% 
  rename(Maximum = maximum)

#df <- eck_stats
#rownames(df)[rownames(df) == "frond_mass"] <- "Frond mass"
#
#  eck_stats <- eck_stats %>% 
#  column_to_rownames(var = "Morphology") %>% 
#  mutate(Morphology = replace(Morphology, 
#                              Morphology == "epi_length", "Epiphyte length (cm)")) %>% 
#  mutate(Morphology = replace(Morphology, 
#                              Morphology == "frond_length", "Frond length (cm)")) %>% 
#  mutate(Morphology = replace(Morphology, 
#                              Morphology == "frond_mass", "Frond mass (kg)")) %>% 
#  mutate(Morphology = replace(Morphology, 
#                              Morphology == "primary_length", "Primary length (cm)")) %>% 
#  mutate(Morphology = replace(Morphology, 
#                              Morphology == "primary_width", "Primary width (cm)")) %>% 
#  mutate(Morphology = replace(Morphology, 
#                              Morphology == "stipe_circ", "Stipe circumference (mm)")) %>% 
#  mutate(Morphology = replace(Morphology, 
#                              Morphology == "stipe_length", "Stipe length (cm)")) %>% 
#  mutate(Morphology = replace(Morphology, 
#                              Morphology == "stipe_mass", "Stipe mass (kg)")) %>% 
#  mutate(Morphology = replace(Morphology, 
#                              Morphology == "total_length", "Total length (cm)")) %>% 
#  mutate(Morphology = replace(Morphology, 
#                              Morphology == "tufts", "Number of tufts"))

@@

```{r echo=FALSE, message=FALSE, warning=FALSE}

library(knitr)
library(kableExtra)

knitr::kable(eck_stats, format = "html") %>%
  kable_styling()

Insert table of spherecity factors.

>>>>

---
title: 'Numerical trajectory patterns of floating macroalgae'
author: "Ross Coppin"
output:
  html_document:
  pdf_document:
    fig_caption: yes
  word_document:
bibliography: references.bib
csl: markdown/frontiers.csl.txt
---

@@

To accurately determine the trajectory of marine macroalgae these aspects need to be taken into account. Past research by @allen1999 and @breivik2011 have provided estimates of drag for various objects based on experimental work, which consisted of both direct and indirect methods of the Leeway model. However, this work does not consider biological material such as macroalgae. Although drag estimates do exist, these have been calculated for macroalgae not detached from the substratum and are regarded as fixed-point estimates. Methods used in the field of chemical engineering provide an alternative approach to estimating drag of macroalgae by means of formulas to calculate drag based on shape. Chemical engineers in the field of fluid dynamics which come across similar problems when including drag use an approach known as a 'spherecity factor'. A spherecity factor is expressed mathematically by $\psi=s/s_d$, and the ratio of the surface area of a sphere having the volume as the irregular shaped particle ($s$) to the actual surface area of the particle ($s_d$). The values are dependent on the shape of the object and a range of values have been determined based on numerous experimental work.

In this study, a hypothetical approach was used to shed light on the role of drag on macroalgal drift trajectory by means of Langrangian based trajectory simulations. These simulations aimed to determine the role that hydrodynamic and wind drag might play in drift trajectory of macroalgae. In terms of wind drag, the direct effect of wind on the kelp was estimated rather than its affect on surface-currents.

@@

## Load software need for entire script

library(ggplot2)
library(rgdal)
library(tidyverse)
library(viridis)
library(gridExtra)
library(ggpubr)
#library(ggord)
library(vegan)
library(dplyr)
library(ggpubr)
library(maps)

source("map/scaleBarFunc.R")

@@

@@

This study simulated two scenarios; one where the plant is fully submerged and only exposed to seawater drag; and the other where parts of the plants are partially exposed resulting in a combination of seawater and air form drag. Drag based on shape based coefficients were used in the calculations for determining the overall kelp velocity vectors. A momentum energy equation was used to calculate the drag forces for the relevant simulation. This approach has been employed in modeling iceberg drift trajectory [@andersson2017; @lichey2001; @eik2009] and was adapted to suit this particular study. The energy momentum equation used to calculate the drag force exerted on the virtual kelp particle was, $$m({d_u/d_t})=F_d + F_a + F_f$$
where $m$ is kelp mass, $d_u$ the virtual kelp particle velocity, $F_d$ the hydrodynamic drag, $F_a$ is the wind drag, and $F_f$ the surface-current flow. The assumption made was that the mass of the kelp did not change for each virtual kelp particle over the course of the simulations. Hydrodynamic drag, wind drag and surface current-flow are two-dimensional vector quantities and were calculated for each time step of each simulation.

@@

In order to incorporate hydrodynamic and wind drag, the cross-sectional area of the kelp was calculated first. Known geometric shapes reflecting the relevant plant sections were used to estimate the surface area for various parts of the plant, for details please see image/table ???. The dimensional data needed was estimated in cases were data was not available for that particular morphological characteristic. The bulb/pneumatocyst is a highly variable morphological characteristic and in some cases can appear absent, the same is true for the holdfast. Therefore, the cross-sectional area of the bulb/pneumatocyst was not considered in the calculation of overall cross-sectional area. In terms of the holdfast, a standard cross-sectional area was used for simulations where the inclusion of the holdfast was considered. The cross-sectional area calculated was site specific and the dimensions needed were garnered from morphology data from a previous study [@coppin2020]. The minimum, maximum and mean were calculated for each morphological characteristic (see appendix), which were used for calculating the overall minimum, maximum and mean cross-sectional areas needed to run the various simulations. The trajectory of these "types" of cross-sectional areas (minimum, maximum and mean overall cross-sectional area) were used to determine the influence of drag, both hydrodynamic and wind, on overall trajectory.

```{r sim table, echo=FALSE, message=FALSE, warning=FALSE}

library(flextable)
library(officer)

sim_table <- read.csv("tables/chapter_3/simulation_table.csv", sep = ";")

sim_table <- flextable(data = sim_table)
sim_table <- theme_zebra(sim_table, odd_header = "transparent", even_header = "transparent")
sim_table <- set_header_labels(sim_table, Particle.behaviour = "Particle behaviour", Cross.sectional.area = "Cross-sectional area", Time.Range = "Time Range", Release.Lon = "Release longitude", Release.Lat = "Release Latitude", Hydrodynamic.model = "Hydrodynamic model", Wind.Model = "Wind model")
sim_table <- align(x= sim_table, align = "center", part = "all")
sim_table

@@

```{r morphology_wrangling, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
# Investigate morphology data to assist in estimating surface area
# Load software
library(tidyverse)
library(ggpubr)
library(gridExtra)
library(ggplot2)
library(vegan)
library(dplyr)
# Read in morphology data
eck_df <- read.csv("data/morph_data/eck_morph_data.csv", sep = ";") %>%
  filter(site!= "Rocky Bank") %>%   # Not using Rocky Bank in this study
  filter(site!= "De Hoop") 
eck_df$site <- as.character(eck_df$site)
eck_df <- eck_df %>% 
  filter(population == "Deep") 
eck_long_df <- gather(eck_df, Morphology, measurement, frond_mass:total_length, factor_key=TRUE)
# Summary stats to make it easier to refer to later, these will not be used any analysis.
site_list <- c("Kommetjie")
eck_stats <- eck_long_df %>% 
  group_by(site, Morphology) %>% 
  summarise(mean = mean(measurement, na.rm = T),
            minimum = min(measurement, na.rm = T),
            maximum = max(measurement, na.rm = T)) %>% 
  ungroup() %>% 
  filter(site %in% site_list) %>% 
  rename(Site = site) %>% 
  rename(Mean = mean) %>% 
  rename(Minimum = minimum) %>% 
  rename(Maximum = maximum)

eck_stats <- flextable(data = eck_stats)
eck_stats <- theme_zebra(eck_stats, odd_header = "transparent", even_header = "transparent")
eck_stats <- align(x= eck_stats, align = "center", part = "all")
eck_stats

@@
